Type-Safe Modular Parsing

### 2017-01-14 paper/sections/Discussion.tex
Section 5

The use of inheritance-based approach and Object Algebras enables us to build modular parsers, which are able to evolve with syntax together. This section explores more interesting features, including parsing multi-sorted syntax, overriding existing parsing rules, language components for abstracting language features, and alternative techniques under the whole framework.

\subsection{Parsing Multi-sorted Syntax}\label{subsec:differentsyntax}

@@

Another advantage of using Object Algebras is that it is easy to
model multi-sorted languages. If the syntax contains multiple sorts, we can distinguish them by different type parameters. As an example, we extend the language of literals, additions and variables by introducing types as a new kind of type argument, and we add lambda abstractions to expressions. Now the language has two sorts in the syntax, types and expressions:\\

@@

As many syntactically extensible parsers, our approach also supports changing part of existing parsers, but in a type-safe way. We can easily update existing parsing rules with new implementations, or eliminate them in extended parsers. As an illustration, suppose we have an untyped lambda abstraction case in a base parser as below. \inlinecode{pLam} parses a lambda symbol, an identifier, a dot and an expression in sequence. Some irrelevant details are omitted.

@@

Then we want to replace the untyped lambda abstractions by typed lambdas. With inheritance and method overriding, it is easy to only change the implementation of \lstinline{pLam} in the extended parser. The new \inlinecode{pLam} parses a colon and a type in addition, between the identifier and the dot. Due to dynamic dispatch, our new explanation of lambdas will make a difference without affecting the other parts of the parser.

@@

\subsection{Language Component}\label{subsec:language-component}

@@

Instead of design and build a language from scratch, we can easily add a new feature by reusing the corresponding language component. For example, if a language is composed with a component of boolean expressions, including if-then-else, it immediately knows how to parse, traverse, and pretty-print if-then-else structure! It could be very useful for rapid developing DSLs.

@@

Such composition shows perfect modularity. The only
drawback is that the glue code of composition appears to be boilerplate. As shown above, we are combining ASTs, parsers and pretty-printers of
\lstinline{VarExpr} and \lstinline{TypedLam} respectively. Such a pattern refers to \textit{family polymorphism}~\cite{ernst01FP} which is unfortunately not supported in Scala. Nonetheless, one can avoid such boilerplate by using metaprogramming techniques.

@@

Under our modular parsing framework, we use Packrat parsing as the underlying parsing technique, OO inheritance with method overriding for composing and extending parsers, and Object Algebras for parsing extensible ASTs. However, our framework is more general and thus more powerful, because those aspects are orthogonal to each other, and hence can have their alternatives.

@@

Our approach does not depend on a particular parser combinator library or parsing algorithm. We demonstrated that Packrat parsing has some advantages and is suitable in a modular setting. However, under certain circumstances or in other programming languages, it may not be the best choice. One can use other parser combinator libraries as long as it meets his requirements.

@@

We use traits to model parsers, OO inheritance to compose them, and method overriding for their extensibility. These language features may not be supported in other programming languages, especially in functional languages such as Haskell. Nevertheless, \textit{open recursion} could be used as an alternative. An explicit ``self-reference'' parameter is able to explain the recursive calls dynamically, by the real argument passed at runtime.

@@

Besides Object Algebras, many other techniques including \textit{Data types à la carte} (DTC)~\cite{swierstra2008data} and \textit{Modular reifiable matching} (MRM)~\cite{Oliveira2015} also support extensible data structures. These alternatives can also be adopted to build ASTs and corresponding parsers.

@@

As we demonstrated, the whole modular parsing framework itself is modular, which means it can be customized easily. Actually, we conducted some previous experiments of modular paring in Haskell, based on open recursion, MRM and Parsec. However, we suffered from issues of Parsec, as illustrated in Section~\ref{subsec:challenges}.

>>>>

The use of inheritance-based approach and Object Algebras enables us
to build modular parsers, which are able to evolve with together with
syntax. This section explores more interesting features, including
parsing multi-sorted syntax, overriding existing parsing rules,
language components for abstracting language features, and alternative
techniques under the whole framework.

\subsection{Parsing Multi-Sorted Syntax}\label{subsec:differentsyntax}

@@

Using Object Algebras is that it is easy to
model multi-sorted languages. If the syntax contains multiple sorts, we can distinguish them by different type parameters. As an example, we extend the language of literals, additions and variables by introducing types as a new kind of type argument, and we add lambda abstractions to expressions. Now the language has two sorts in the syntax, types and expressions:\\

@@

As many syntactically extensible parsers, our approach also supports
changing part of existing parsers, but in a type-safe way. We can
easily update existing parsing rules with new implementations, or
eliminate them in extended parsers. This can be useful in many
situations, including when conflicts or ambiguities arize when
composing languages.
As an illustration, suppose we have an untyped lambda abstraction case in a base parser as below. \inlinecode{pLam} parses a lambda symbol, an identifier, a dot and an expression in sequence. Some irrelevant details are omitted.

@@

Then we want to replace the untyped lambda abstractions by typed
lambdas. With inheritance and method overriding, it is easy to only
change the implementation of \lstinline{pLam} in the extended parser.
The new \inlinecode{pLam} parses a colon and a type in addition,
between the identifier and the dot. Due to dynamic dispatch, our new
implementation of lambdas will be different without affecting the other parts of the parser.

@@

\subsection{Language Components}\label{subsec:language-component}

@@

Instead of designing and building a language from scratch, we can
easily add a new feature by reusing the corresponding language
component. For example, if a language is composed from a component of
boolean expressions, including if-then-else, it immediately knows how
to parse, traverse, and pretty-print the if-then-else structure.
Grouping language features in this way can
be very useful for rapid development of DSLs.

@@

The only drawback is that the glue code of composition appears to be
boilerplate. As shown above, we are combining ASTs, parsers and
pretty-printers of \lstinline{VarExpr} and \lstinline{TypedLam}
respectively. Such a pattern refers to \textit{family
  polymorphism}~\cite{ernst01FP} which is unfortunately not fully supported
in Scala, since nested classes/traits have to be manually composed. 
%Nonetheless, one can avoid such boilerplate by using metaprogramming techniques.

@@

Under our modular parsing framework, we use Packrat parsing as the underlying parsing technique, OO inheritance with method overriding for composing and extending parsers, and Object Algebras for parsing extensible ASTs. However, our framework is more general and thus more powerful, because those aspects are orthogonal to each other, and hence can have alternatives.

@@

Our approach does not depend on a particular parser combinator library
or parsing algorithm. We demonstrated that Packrat parsing has some
advantages and is suitable in a modular setting. However, under
certain circumstances or in other programming languages, it may not be
the best choice. One can use other parser combinator libraries as long
as th guidelines proposed in Section~\ref{sec:packrat} are met.

@@

  We use traits to model parsers, OO inheritance to compose them, and
  method overriding for their extensibility. These language features
  may not be supported in other programming languages, especially in
  functional languages such as Haskell. Nevertheless, \textit{open
    recursion}~\cite{CookThesis} could be used as an alternative. An explicit
  ``self-reference'' parameter is able to explain the recursive calls
  dynamically, by the real argument passed at runtime.

@@

Besides Object Algebras, many other techniques including \textit{Data
  types à la carte} (DTC)~\cite{swierstra2008data} ot the Cake pattern~\cite{odersky2005independently} also support extensible data structures. These alternatives could also be adopted to build ASTs and corresponding parsers.

@@

As we demonstrated, the whole modular parsing framework itself is
modular, which means it can be customized easily. 
%We conducted some previous experiments of modular parsing in Haskell, based on open recursion, MRM and Parsec. However, we suffered from issues of Parsec, as illustrated in Section~\ref{subsec:challenges}.
